humanCdGF_predict_cfu

### 2016-11-22 code/HumanCdGF_predict_cfu.Rmd
remove code reducing features of 1%  rel abun
remove initial model feature reductions

# remove samples not used or with other cdiff strains
meta_file <- 
  meta_data %>% 
  filter(Early_Euth == FALSE, cdiff_strain == 431, !cage_id == 'NP1') %>% 
  select(mouse_id, sample_id, day, log_cfu) %>% 
  filter(!is.na(log_cfu))

shared_list_days <- select(meta_file, -log_cfu)

# create shared with relative abundances over 1%
mean_otu_counts <- unique(apply(shared_file, 1, sum))
rel_abund_shared <- 100 * shared_file / mean_otu_counts
rel_abund_shared_01 <- rel_abund_shared[ , apply(rel_abund_shared, 2, max) > 1]
# remove correlative otus
rel_abund_cor <- cor(rel_abund_shared_01, method = 'spearman')
rownames(rel_abund_cor) <- colnames(rel_abund_shared_01) # add row names to identify otus correlated with features
cor_otus <- findCorrelation(rel_abund_cor, cutoff = cor_cutoff)
rel_abund_01_filtr <- rel_abund_shared_01[,-cor_otus]
rel_abund_01_filtr$sample_id <- rownames(rel_abund_01_filtr)
```

@@

```{r predict median CFU, fig.width=10,fig.height=10}

colonized_cfu <- meta_data %>% 
  filter(day %in% c(4:10), !cage_id == 'NP1', 
         Early_Euth == FALSE, cdiff_strain == 431) %>% 
  select(mouse_id, cage_id, log_cfu) %>% 
  group_by(mouse_id, cage_id) %>% 
  summarise(log_cfu_med = median(log_cfu)) %>% 
  ungroup()
sample_list_d0 <- shared_list_days %>% 
    filter(day == 0) %>% 
    select(-day)
community_d0.df <- rel_abund_01_filtr %>% 
    right_join(sample_list_d0) %>% 
    select(-sample_id)
cfu_shared_colonized.df <- colonized_cfu %>% 
    right_join(community_d0.df) %>% 
    select(-mouse_id, -cage_id) %>% 
    filter(!is.na(log_cfu_med))

@@

rfe_ctrl <- rfeControl(functions = rfRFE,
                   method = "repeatedcv",
                   repeats = 10,
                   verbose = FALSE)
subsets <- c(1:5, c(seq(10, 70, 10)))
set.seed(seed)
rf_cfu_model <- rfe(cfu_shared_colonized.df[ , -1], cfu_shared_colonized.df$log_cfu_med,
                 sizes = subsets,
                 rfeControl = rfe_ctrl)
top_important_OTU <- predictors(rf_cfu_model)

@@

#One per cage
one_mouse_initial_df <- data.frame(colonized_cfu %>% 
    right_join(community_d0.df) %>% 
    filter(!is.na(log_cfu_med))  %>% 
    select(log_cfu_med, mouse_id, cage_id, one_of(top_important_OTU)))

om_raw_otus <- data.frame(otu = colnames(select(one_mouse_initial_df, contains('Otu'))))
for(i in 1:iters){
    set.seed(i)
    train <- one_mouse_initial_df %>% 
      group_by(cage_id) %>% 
      sample_n(1) %>% ungroup() 
    test <- one_mouse_initial_df %>% 
      filter(!mouse_id %in% train$mouse_id)
    set.seed(seed)
    rf_om <- randomForest(log_cfu_med ~ ., data = select(train, -mouse_id, -cage_id),
                             importance = TRUE, keep.forest = T, ntree = n_trees)
    import_df <- importance(rf_om)
    colnames(import_df)[1] <- paste(train$mouse_id, collapse = ' ')
    om_raw_otus <- merge(om_raw_otus, import_df, by.x = 'otu', by.y = 'row.names')
    om_raw_otus <- select(om_raw_otus, -IncNodePurity)
}
raw_importance <- data.frame(om_raw_otus[ , -1], row.names = om_raw_otus$otu)
### rank importance
importance_rank <- apply(raw_importance, 2, function(x) rank(x, na.last = F))
raw_rank <- apply(importance_rank, 1, function(x) sum(x, na.rm = T))
final_rank <- sort(raw_rank[raw_rank >=  ncol(importance_rank)*sum(importance_rank[ , 1])/nrow(importance_rank)], decreasing = T)

one_mouse_df <- data.frame(colonized_cfu %>% 
    right_join(community_d0.df) %>% 
    filter(!is.na(log_cfu_med))  %>% 
    select(log_cfu_med, mouse_id, cage_id, one_of(names(final_rank))))

@@

layout(matrix(c(1:9), ncol=3, byrow=TRUE))
for(otu in seq_along(importance_om$tax)){
    partialPlot(rf_n_features, data.frame(cfu_shared_colonized.df), x.var=rownames(importance_om)[otu],
                xlab='', main = NULL)
    if(length(features_cor[[otu]]) > 1){
      title(main = paste0('Unidentifiable Feature (', 
                          get_tax(row_list = rownames(importance_om)[otu])$tax_label,
                          ' Shown)'))
      mtext(side = 1, line = 3, cex = 0.6,
        paste('Correlated features: \n', 
              paste(get_tax(row_list = features_cor[[otu]])$tax_label, collapse = ', ')))
      } else {
      title(paste(get_tax(row_list = rownames(importance_om)[otu])$tax_label))
        }
    par(new = T)
    plot(y = partial_df[ , 'log_cfu_med'],
         x = partial_df[ , rownames(importance_om)[otu]], 
         pch = NA, bg = 'gray', xlab = NA, ylab = NA, axes = F)
    axis(side = 4)
    mtext(side = 4, line = 3, 'Log10 CFU')
# To color points by mouse - Change pch to 21 and un-comment following lines
  lapply(seq_along(unique(partial_df$cage_id)), function(m){
    mouse <- unique(partial_df$cage_id)[m]
    points(y = partial_df[partial_df$cage_id %in% mouse, 'log_cfu_med'],
           x = partial_df[partial_df$cage_id %in% mouse , rownames(importance_om)[otu]], 
           pch = 21, cex = 2, bg = color[m], xlab = NA, ylab = NA, axes = F)
  })
}
```

@@

mouse_list <- meta_data %>% 
  filter(!is.na(day), !cage_id == 'NP1',  
         Early_Euth == FALSE, cdiff_strain == 431) %>% 
  select(mouse_id) %>% 
  unique() %>% 
  filter(!mouse_id == 'NP2_983') # remove NP2_983 no colonization data

@@

#One per cage
one_mouse_cr_df <- LOcO_cr_df

om_raw_otus_cr <- data.frame(otu = colnames(select(one_mouse_cr_df, contains('Otu'))))
for(i in 1:iters){
    set.seed(i)
    train <- one_mouse_cr_df %>% 
      group_by(cage_id) %>% 
      sample_n(1) %>% ungroup() 
    test <- one_mouse_cr_df %>% 
      filter(!mouse_id %in% train$mouse_id)
    set.seed(seed)
    rf_om <- randomForest(ratio ~ ., data = select(train, -mouse_id, -cage_id),
                             importance = TRUE, keep.forest = T, ntree = n_trees)
    import_df <- importance(rf_om)
    colnames(import_df)[1] <- paste(train$mouse_id, collapse = ' ')
    om_raw_otus_cr <- merge(om_raw_otus_cr, import_df, by.x = 'otu', by.y = 'row.names')
    om_raw_otus_cr <- select(om_raw_otus_cr, -IncNodePurity)
}
raw_importance_cr <- data.frame(om_raw_otus_cr[ , -1], row.names = om_raw_otus_cr$otu)
### rank importance
importance_rank_cr <- apply(raw_importance_cr, 2, function(x) rank(x, na.last = F))
raw_rank_cr <- apply(importance_rank_cr, 1, function(x) sum(x, na.rm = T))
final_rank_cr <- sort(raw_rank_cr[raw_rank_cr >=  ncol(importance_rank_cr)*sum(importance_rank_cr[ , 1])/nrow(importance_rank_cr)], decreasing = T)

om_cr_df <- data.frame(shared_col_rate_df %>% 
    select(ratio, mouse_id, cage_id, one_of(names(final_rank_cr))))

@@

cage_color <- rainbow(length(unique(cr_prediction$cage_id)))
for (otu in seq_along(cr_features)) {
    partialPlot(rf_col_rate, shared_cr_df, cr_features[otu],
                xlab='', main = NULL)
    if(length(cr_features_cor[[otu]]) > 1){
      title(main = paste0('Unidentifiable Feature (', 
                          get_tax(row_list = cr_features[otu])$tax_label,
                          ' Shown)'))
      mtext(side = 1, line = 3, cex = 0.6,
        paste('Correlated features: \n', 
              paste(get_tax(row_list = cr_features_cor[[otu]])$tax_label, collapse = ', ')))
      } else {
      title(paste(get_tax(row_list = cr_features[otu])$tax_label))
        }
  par(new = T)
    plot(y = shared_col_rate_df[ , 'ratio'],
         x = shared_col_rate_df[ , cr_features[otu]], 
         pch = NA, bg = 'gray', xlab = NA, ylab = NA, axes = F)
    axis(side = 4)
    mtext(side = 4, line = 3, 'Relative Colonization')
# To color points by mouse - Change pch to 21 and un-comment following lines
  lapply(seq_along(unique(shared_col_rate_df$cage_id)), function(m){
    mouse <- unique(shared_col_rate_df$cage_id)[m]
    points(y = shared_col_rate_df[shared_col_rate_df$cage_id %in% mouse, 'ratio'],
           x = shared_col_rate_df[shared_col_rate_df$cage_id %in% mouse , cr_features[otu]], 
           pch = 21, cex = 2, bg = color[m], xlab = NA, ylab = NA, axes = F)
  })
}
```

>>>>

# remove samples not used or with other cdiff strains
meta_data <- 
  meta_data %>% 
  filter(Early_Euth == FALSE, cdiff_strain == 431, !cage_id == 'NP1') %>% 
  select(mouse_id, sample_id, cage_id, day, log_cfu) %>% 
  filter(!is.na(log_cfu))
shared_list_days <- meta_data %>% 
  filter(day == 0) %>% 
  select(-day, -log_cfu, -cage_id)
shared_file <- shared_file[rownames(shared_file) %in% shared_list_days$sample_id, ]

# create shared with relative abundances over 1%
mean_otu_counts <- unique(apply(shared_file, 1, sum))
rel_abund_shared <- shared_file
rel_abund_shared_01 <- rel_abund_shared[ , apply(rel_abund_shared, 2, max) > 1]
# remove correlative otus
rel_abund_cor <- cor(rel_abund_shared_01, method = 'spearman')
rownames(rel_abund_cor) <- colnames(rel_abund_shared_01) # add row names to identify otus correlated with features
cor_otus <- findCorrelation(rel_abund_cor, cutoff = cor_cutoff)
rel_abund_01_filtr <- rel_abund_shared_01[,-cor_otus]
community_d0.df <- merge(shared_list_days, rel_abund_01_filtr, 
                         by.x = 'sample_id', by.y = 'row.names') %>% 
                    select(-sample_id)
```

@@

```{r predict median CFU, fig.width=10,fig.height=10}
colonized_cfu <- meta_data %>% 
  filter(day %in% c(4:10)) %>% 
  group_by(mouse_id, cage_id) %>% 
  summarise(log_cfu_med = median(log_cfu)) %>% 
  ungroup()

cfu_shared_colonized.df <- colonized_cfu %>% 
    right_join(community_d0.df) %>% 
    select(-mouse_id, -cage_id) %>% 
    filter(!is.na(log_cfu_med))

@@

rfe_ctrl <- rfeControl(functions = rfRFE,
                   method = "repeatedcv",
                   repeats = 10,
                   verbose = FALSE)
subsets <- c(1:20, c(seq(25, 85, 10)))
set.seed(seed)
rf_cfu_model <- rfe(cfu_shared_colonized.df[ , -1], cfu_shared_colonized.df$log_cfu_med,
                 sizes = subsets,
                 rfeControl = rfe_ctrl)
top_important_OTU <- predictors(rf_cfu_model)

@@

#One per cage
one_mouse_df <- data.frame(colonized_cfu %>% 
    right_join(community_d0.df) %>% 
    filter(!is.na(log_cfu_med))  %>% 
    select(log_cfu_med, mouse_id, cage_id, one_of(top_important_OTU)))

@@

layout(matrix(c(1:9), ncol=3, byrow=TRUE))
for(otu in seq_along(importance_om$tax)){
    partialPlot(rf_n_features, data.frame(cfu_shared_colonized.df), x.var=rownames(importance_om)[otu],
                xlab='', main = NULL)
    if(length(features_cor[[otu]]) > 1){
      title(main = paste0('Unidentifiable Feature\n(', 
                          get_tax(row_list = rownames(importance_om)[otu])$tax_label,
                          ' Shown)'))
      mtext(side = 1, line = 3, cex = 0.6,
        paste('Correlated features: \n', 
              paste(get_tax(row_list = features_cor[[otu]])$tax_label, collapse = ', ')))
      } else {
      title(paste(get_tax(row_list = rownames(importance_om)[otu])$tax_label))
        }
    par(new = T)
    plot(y = partial_df[ , 'log_cfu_med'],
         x = partial_df[ , rownames(importance_om)[otu]], 
         pch = NA, bg = 'gray', xlab = NA, ylab = NA, axes = F)
    axis(side = 4)
    mtext(side = 4, line = 3, 'Log10 CFU')
# To color points by mouse - Change pch to 21 and un-comment following lines
  lapply(seq_along(unique(partial_df$cage_id)), function(m){
    mouse <- unique(partial_df$cage_id)[m]
    points(y = partial_df[partial_df$cage_id %in% mouse, 'log_cfu_med'],
           x = partial_df[partial_df$cage_id %in% mouse , rownames(importance_om)[otu]], 
           pch = 21, cex = 2, bg = color[m], xlab = NA, ylab = NA, axes = F)
  })
}
```

@@

mouse_list <- meta_data %>% 
  select(mouse_id) %>% 
  unique() %>% 
  filter(!mouse_id == 'NP2_983') # remove NP2_983 no colonization data

@@

#One per cage
om_cr_df <- LOcO_cr_df

@@

cage_color <- rainbow(length(unique(cr_prediction$cage_id)))
for (otu in seq_along(cr_features)) {
    partialPlot(rf_col_rate, shared_cr_df, cr_features[otu],
                xlab='', main = NULL)
    if(length(cr_features_cor[[otu]]) > 1){
      title(main = paste0('Unidentifiable Feature\n(', 
                          get_tax(row_list = cr_features[otu])$tax_label,
                          ' Shown)'))
      mtext(side = 1, line = 3, cex = 0.6,
        paste('Correlated features: \n', 
              paste(get_tax(row_list = cr_features_cor[[otu]])$tax_label, collapse = ', ')))
      } else {
      title(paste(get_tax(row_list = cr_features[otu])$tax_label))
        }
  par(new = T)
    plot(y = shared_col_rate_df[ , 'ratio'],
         x = shared_col_rate_df[ , cr_features[otu]], 
         pch = NA, bg = 'gray', xlab = NA, ylab = NA, axes = F)
    axis(side = 4)
    mtext(side = 4, line = 3, 'Relative Colonization')
# To color points by mouse - Change pch to 21 and un-comment following lines
  lapply(seq_along(unique(shared_col_rate_df$cage_id)), function(m){
    mouse <- unique(shared_col_rate_df$cage_id)[m]
    points(y = shared_col_rate_df[shared_col_rate_df$cage_id %in% mouse, 'ratio'],
           x = shared_col_rate_df[shared_col_rate_df$cage_id %in% mouse , cr_features[otu]], 
           pch = 21, cex = 2, bg = color[m], xlab = NA, ylab = NA, axes = F)
  })
}
```
